Automatic prostate cancer detection through \acs*{dce}-\acs*{mri} images: all you need is a good normalization

### 2017-01-27 03_experiments/experiments.tex
make english correction

The multi-parametric \ac{mri} data are acquired from a cohort of patients with higher-than-normal level of \ac{psa}.
The acquisition is performed using a 3T whole body \ac{mri} scanner (Siemens Magnetom Trio TIM, Erlangen, Germany) using sequences to obtain \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri} and \ac{dw}-\ac{mri}.
Aside of the \ac{mri} examination, these patients also have undergone a guided-biopsy.
The dataset is composed of a total of 20 patients of which 18 patients have biopsy proven \ac{cap} and 2 patients are ``healthy'' with negative biopsies.
Therefore, 13 patients have a \ac{cap} in the \ac{pz}, 3 patients have \ac{cap} in the \ac{cg}, 2 patients have invasive \ac{cap} in both \ac{pz} and \ac{cg} and finally 2 patients are considered as ``healthy''.
An experienced radiologist has segmented the prostate organ --- on \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} --- as well as the prostate zones (i.e., \ac{pz} and \ac{cg}) and \ac{cap} on the \ac{t2w}-\ac{mri}.

A \SI{3}{\mm} slice fat-suppressed \ac{t2w} fast spin-echo sequence (\ac{tr}/\ac{te}/\ac{etl}: \SI{3400}{\ms}/\SI{85}{\ms}/13) is used to acquire images in sagittal and oblique coronal planes, the latter planes being orientated perpendicular or parallel to the prostate \ac{pz} – rectal wall axis.
Three-dimensional \ac{t2w} fast spin-echo (\ac{tr}/\ac{te}/\ac{etl}: \SI{3600}{\ms}/\SI{143}{\ms}/109, slice thickness: \SI{1.25}{\mm}) images are then acquired in an oblique axial plane.
The nominal matrix and \ac{fov} of the 3D \ac{t2w} fast spin-echo images are $320 \times 256$ and $280 \times 240$ mm\textsuperscript{2}, respectively, thereby affording sub-millimetric pixel resolution within the imaging plane.

\ac{dce}-\ac{mri} is performed using a fat suppressed 3D T$_1$ VIBE sequence (\ac{tr}/\ac{te}/Flip angle: \SI{3.25}{\ms}/\SI{1.12}{\ms}/\SI{10}{\degree}; Matrix: $256 \times 192$; \ac{fov}: $280 \times 210$ (with 75\% rectangular \ac{fov}); slab of 16 partitions of \SI{3.5}{\mm} thickness; temporal resolution: \SI{6}{\s}/slab over approximately \SI{5}{\minute}).
A power injector (Medrad, Indianola, USA) is used to provide a bolus injection of Gd-DTPA (Dotarem, Guerbet, Roissy, France) at a dose of \SI{0.2}{\ml} Gd-DTPA/kg of body weight.

These \ac{dce}-\ac{mri} sequences are resampled using the spatial information of the \ac{t2w}-\ac{mri} and missing data are interpolated using a linear interpolation.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered, to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} in order to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\citep{ibanez2005itk}.

@@

The implementation of the registration (C++), normalization (Python), and classification pipeline (Python) are publicly available on GitHub\footnote{\url{https://github.com/I2Cvb/lemaitre-2016-nov/tree/master}}~\citep{lemaitre2016github}.
The data used for this work are also publicly available\footnote{\url{https://zenodo.org/record/61163}}~\citep{lemaitre2016dce}.

@@

\begin{table*}
  \caption{Coefficient of determination $R^{2}$ (i.e., $\mu \ (\pm \sigma)$), while fitting data with the different quantification models.}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tabular}{lc c c c c c}
    \toprule
    Data type & Brix & Hoffmann & Tofts population \ac{aif} & Tofts patient \ac{aif} & \ac{pun} & Semi-quantitative \\
    \midrule
    Un-normalized & $0.85 \ (\pm 0.11)$ & $0.81 \ (\pm 0.17)$ & $0.84 \ (\pm 0.14)$ & $0.88 \ (\pm 0.12)$ & $0.27 \ (\pm 0.18)$ & $0.64 \ (\pm 0.24)$  \\
    Normalized    & $0.92 \ (\pm 0.05)$ & $0.72 \ (\pm 0.32)$ & $0.92 \ (\pm 0.06)$ & $0.90 \ (\pm 0.10)$ & $0.28 \ (\pm 0.20)$ & $0.75 \ (\pm 0.20)$  \\
    \bottomrule
  \end{tabular}
  }
  \label{tab:r2}
\end{table*}

Parameter estimation of the quantification methods are related to fit a specific model to the \ac{dce}-\ac{mri} data.
Therefore, this section reports the goodness of fitting by computing the coefficient of determination $R^2$ such as in Eq.\,\eqref{eq:r2}

@@

\noindent where $s_t$ and $\hat{s}_t$ are the signal to be fitted and the estimated signal at time $t$, respectively; $\bar{s}$ is the average signal to be fitted.

Mean and standard-deviation of the coefficient of determination $R^{2}$ is reported in Table~\ref{tab:r2} for each quantification model.
Brix, Hoffmann, and Tofts models are fitted with a coefficient $R^{2}$ superior to 0.80.
Additionally, the proposed \ac{pun} model does not to fit well the data.
After introspection of the fitted curves, the original model --- as
formulated in Eq.\,\eqref{eq:pun} --- does not provide enough degrees
of freedom to fit well the data. Additional parameters should be
integrated in this model to control the translation and amplitude of
the model to be fitted.
Data normalization improves the coefficient $R^2$ for all the methods apart of the Hoffmann model.
The large standard deviation for this model might imply that there are some cases where the fitting fails.

@@

To study the potential benefit of our normalization, \ac{cap} are detected at a voxel level using pharmacokinetic parameters estimated from un-normalized and normalized \ac{dce}-\ac{mri} data.
Each individual pharmacokinetic parameter is classified to evaluate their individual discriminative power to detect \ac{cap}.
Therefore, a \ac{rf} classifier is used in conjunction with a \ac{lopo}.
The use of \ac{rf} is motivated since that it leads to the best performance in the state-of-the-art methods~\citep{litjens2014computer,lemaitre2015computer}.
Results are summarized in Table~\ref{tab:resfeats} in terms of \ac{auc}.
Normalization can improve the detection of \ac{cap}; however, the benefit of normalization is more obvious by combining together the pharmacokinetic features of a given model (e.g., $A$, $k_ep$, and $k_el$ for Brix model), as previously done in traditional \ac{cad} system~\citep{lemaitre2015computer}.
For the latter configuration, results are summarized by performing a \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
Quantification using normalized data outperforms quantification using un-normalized data in terms of classification performance apart of Hoffmann and Tofts population-based \ac{aif} models.
The reasons behind the decrease of the \ac{auc} might be related to: (i) a poor fitting as discussed in Sect.\,\ref{sec:fit} (cf., Hoffmann model) and (ii) a small number of patients while estimating some parameters (cf., Tofts model).
The best classification performance are obtained using the semi-quantitative approach with an \ac{auc} of $0.655$.

@@

As stated in the introduction, the quantification methods are extracting a set of parameters characterizing the enhancement \ac{dce}-\ac{mri} signal.
However, this extraction might lead to a loss of information.
This experiment is performed to assess if making use of the whole \ac{dce}-\ac{mri} signal instead of the just the pharmacokinetic parameters can improve the classification performance.
Therefore, each enhanced \ac{dce}-\ac{mri} signal, normalized and un-normalized, is classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The \ac{roc} analysis and \ac{auc} are reported in Fig.\,\ref{fig:rfnormdcesignal}.
Classification without normalization lead to the worst performance, with an \ac{auc} of $0.568$.
However, data normalization in conjunction with the use of the entire \ac{dce}-\ac{mri} signal is the strategy which outperforms others, with an \ac{auc} of $0.666$.
Comparable outcomes are visible while analyzing the \ac{auc}
distributions scores for each prostate zone. The \ac{auc} score
related to \ac{cap} in \ac{cg} improves using normalized data, for
most of the models.

>>>>

The multi-parametric \ac{mri} data are acquired from a cohort of patients with higher-than-normal levels of \ac{psa}.
Acquisition is achieved with a 3T whole body \ac{mri} scanner (Siemens Magnetom Trio TIM, Erlangen, Germany) using sequences to obtain \ac{t2w}-\ac{mri}, \ac{dce}-\ac{mri} and \ac{dw}-\ac{mri}.
In addition to the \ac{mri} examination, these patients also have
undergone a \ac{trus} guided-biopsy.
The dataset is composed of a total of 20 patients, 18 of which have
biopsies that were positive for \ac{cap} and 2 patients are considered
``healthy'' because they have negative biopsies.
Therefore, 13 patients have \ac{cap} in the \ac{pz}, 3 patients have
\ac{cap} in the \ac{cg}, 2 patients have invasive \ac{cap} in both the
\ac{pz} and the \ac{cg}, and finally 2 patients are considered ``healthy''.
An experienced radiologist segmented the prostate organ---on
\ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri}---as well as the prostate
zones (i.e., \ac{pz} and \ac{cg}) and the \ac{cap} on the \ac{t2w}-\ac{mri}.

A \SI{3}{\mm} slice of fat-suppressed \ac{t2w} fast spin-echo sequence
(\ac{tr}/\ac{te}/\ac{etl}: \SI{3400}{\ms}/\SI{85}{\ms}/13) is used to
acquire images in the sagittal and oblique coronal planes, with the latter planes orientated perpendicular or parallel to the prostate \ac{pz} – rectal wall axis.
Three-dimensional \ac{t2w} fast spin-echo (\ac{tr}/\ac{te}/\ac{etl}: \SI{3600}{\ms}/\SI{143}{\ms}/109, slice thickness: \SI{1.25}{\mm}) images are then acquired in an oblique axial plane.
The nominal matrix and the \ac{fov} of the 3D \ac{t2w} fast spin-echo images are $320 \times 256$ and $280 \times 240$ mm\textsuperscript{2}, respectively, thereby affording sub-millimetric pixel resolution within the imaging plane.

\ac{dce}-\ac{mri} is performed using a fat-suppressed 3D T$_1$ VIBE
sequence (\ac{tr}/\ac{te}/Flip angle:
\SI{3.25}{\ms}/\SI{1.12}{\ms}/\SI{10}{\degree}; Matrix: $256 \times
192$; \ac{fov}: $280 \times 210$ (with 75\% rectangular \ac{fov}); a slab of 16 partitions of \SI{3.5}{\mm} thickness; temporal resolution: \SI{6}{\s}/slab over approximately \SI{5}{\minute}).
A power injector (Medrad, Indianola, USA) is used to provide a bolus injection of Gd-DTPA (Dotarem, Guerbet, Roissy, France) at a dose of \SI{0.2}{\ml} Gd-DTPA/kg of body weight.

These \ac{dce}-\ac{mri} sequences are resampled using the spatial
information of the \ac{t2w}-\ac{mri} and any missing data are added using linear interpolation.
The volumes of the \ac{dce}-\ac{mri} dynamic are rigidly registered to remove any patient motion during the acquisition.
Furthermore, a non-rigid registration is performed between the \ac{t2w}-\ac{mri} and \ac{dce}-\ac{mri} to propagate the prostate zones and \ac{cap} ground-truths.
The resampling is implemented in C++ using the Insight Segmentation and Registration Toolkit~\citep{ibanez2005itk}.

@@

The implementation of the registration (C++), normalization (Python), and classification pipeline (Python) are publicly available on GitHub\footnote{\url{https://github.com/I2Cvb/lemaitre-2016-nov/tree/master}}~\citep{lemaitre2016github}.
The data used in this work are also publicly available\footnote{\url{https://zenodo.org/record/61163}}~\citep{lemaitre2016dce}.

@@

\begin{table*}
  \caption{Coefficient of determination $R^{2}$ (i.e., $\mu \ (\pm \sigma)$), while fitting data with the different quantification models.}
  \centering
  \resizebox{\columnwidth}{!}{
  \begin{tabular}{lc c c c c c}
    \toprule
    Data type & Brix & Hoffmann & Tofts population \ac{aif} & Tofts patient \ac{aif} & \ac{pun} & Huisman \\
    \midrule
    Un-normalized & $0.85 \ (\pm 0.11)$ & $0.81 \ (\pm 0.17)$ & $0.84 \ (\pm 0.14)$ & $0.88 \ (\pm 0.12)$ & $0.27 \ (\pm 0.18)$ & $0.64 \ (\pm 0.24)$  \\
    Normalized    & $0.92 \ (\pm 0.05)$ & $0.72 \ (\pm 0.32)$ & $0.92 \ (\pm 0.06)$ & $0.90 \ (\pm 0.10)$ & $0.28 \ (\pm 0.20)$ & $0.75 \ (\pm 0.20)$  \\
    \bottomrule
  \end{tabular}
  }
  \label{tab:r2}
\end{table*}

The parameter estimations from the quantification methods are related to fit a specific model to the \ac{dce}-\ac{mri} data.
Therefore, the coefficient of determination $R^2$ reports the goodness
of fit, as follows:

@@

\noindent where $s_t$ and $\hat{s}_t$ are the signal to be fitted and
the estimated signal at time $t$, respectively, and $\bar{s}$ is the average signal to be fitted.

The mean and the standard-deviation of the coefficient of determination, $R^{2}$, are reported in Table~\ref{tab:r2} for each quantification model.
Brix, Hoffmann, and Tofts models are fitted with a coefficient $R^{2}$ superior to 0.80.
Additionally, the proposed \ac{pun} model does not to fit the data well.
After introspection of the fitted curves, the original model --- as
formulated in Eq.\,\eqref{eq:pun} --- does not provide enough degrees
of freedom to fit well the data. Additional parameters should be
integrated in this model to control the translation and amplitude of
the model to be fitted.
Data normalization improves the coefficient $R^2$ for all the methods
except for the Hoffmann model.
The large standard deviation for this model might imply that there are some cases where the fitting fails.

@@

To study the potential benefit of our normalization, \ac{cap} is detected at a voxel level using pharmacokinetic parameters estimated from non-normalized and normalized \ac{dce}-\ac{mri} data.
Each individual pharmacokinetic parameter is classified to evaluate its individual discriminative power to detect \ac{cap}.
In addition, p-values ($p$) are computed to identify pairs of classifiers
which are significantly different, using a Wilcoxon signed-ranked
test. Those p-values are represented in the color coded co-occurrence
matrix shown in Fig.\,\ref{fig:wilcoxon}, where blue cells correspond
to lower p-values while red cells are greater p-values.
Therefore, a \ac{rf} classifier is used in conjunction with a \ac{lopo}.
The use of \ac{rf} is encouraged since it leads to the best performance in the state-of-the-art methods~\citep{litjens2014computer,lemaitre2015computer}.
The results are summarized in Table~\ref{tab:resfeats} in terms of the \ac{auc}.
Normalization can improve the detection of \ac{cap}, but the benefit
of normalization is more obvious with the combination of the
pharmacokinetic features of a given model (e.g., $A$, $k_ep$, and
$k_el$ for the Brix model) as previously executed in the traditional \ac{cad} system~\citep{lemaitre2015computer}.
For the latter configuration, results are summarized by performing a \ac{roc} analysis and computing the \ac{auc}, as reported in Fig.\,\ref{fig:normpharmarf}.
Quantification using normalized data outperforms quantification using
non-normalized data in terms of the classification performance, except
for Hoffmann and Tofts population-based \ac{aif} models.
The reasons behind the decrease of the \ac{auc} might be related to:
(i) a poor fitting as discussed in Sect.\,\ref{sec:fit} (cf., Hoffmann
model) and (ii) a small number of patients for the estimation of some parameters (cf., Tofts model).
The best classification performance is obtained using the
semi-quantitative approach with an \ac{auc} of $0.655 \pm 0.106$.

@@

As stated in the introduction, the quantification methods extract a
set of parameters that characterize the enhancement of the \ac{dce}-\ac{mri} signal.
However, this extraction might lead to a loss of information.
This experiment is performed to assess whether using the whole
\ac{dce}-\ac{mri} signal, rather than just the pharmacokinetic parameters, can improve the classification performance.
Therefore, each enhanced \ac{dce}-\ac{mri} signal, normalized and un-normalized, is classified using a \ac{rf} classifier in a \ac{lopo} fashion.
The \ac{roc} analysis and \ac{auc} are reported in Fig.\,\ref{fig:rfnormdcesignal}.
Classification without normalization leads to the worst performance,
with an \ac{auc} of $0.568 \pm 0.162$.
However, data normalization in conjunction with the use of the entire
\ac{dce}-\ac{mri} signal is the strategy that outperforms others,
with an \ac{auc} of $0.666 \pm 0.154$.
Comparable outcomes are visible while analyzing the \ac{auc}
scores distribution for each prostate zone as shown in
Fig.\,\ref{fig:boxplot}. The \ac{auc} scores
related to \ac{cap} in \ac{cg} improves using normalized data, for
most of the models.

### 2017-01-27 04_discussions/discussions.tex
make english correction

The experiments conducted in the previous section can give rise to several discussions.
In Tofts quantification, two different approaches have been used to infer the pharmacokinetic parameters: using a population-based or a patient-based \ac{aif}.
The population-based \ac{aif} approach leads to better classification performance.
However, there is two shortcomings to take into account while advancing this fact:
(i) T$_{10}$ parameter has been fixed and not computed from a T$_1$ map and
(ii) the population-based \ac{aif} has been estimated from a cohort of only 17 patients.
These two limitations have to be considered while advancing that population-based \ac{aif} modelling is outperforming patient-based \ac{aif} modelling.

The best classification performance is reached by normalizing the \ac{dce}-\ac{mri} data and use the whole enhanced signal as feature, emphasizing the fact that a loss of information while extracting quantitative parameters.
However, this strategy suffers from one drawback: the training time of the \ac{rf} classifier increases since that from 3 to 5 features, the feature space become a 40 dimensions space.

>>>>

The experiments conducted in the previous section can incite several discussions.
In Tofts quantification, two different approaches have been used to infer the pharmacokinetic parameters: using a population-based or a patient-based \ac{aif}.
The patient-based \ac{aif} approach leads to better classification performance.
However, there are two shortcomings to take into account when considering this fact:
(i) the T$_{10}$ parameter has been fixed and is not computed from a T$_1$ map and
(ii) the population-based \ac{aif} has been estimated from a cohort of only 17 patients.
These two limitations have to be considered when asserting that
population-based \ac{aif} modeling outperforms patient-based \ac{aif} modeling.

The best classification performance is achieved by normalizing the
\ac{dce}-\ac{mri} data and using the entire enhanced signal as a
feature, emphasizing the fact that a \emph{loss of information} may occur while extracting quantitative parameters.
Furthermore, normalization is a less complex process than all
quantification methods and significantly improves the classification
performance of the semi-quantitative approach ($p=0.001$) proposed by
\citeauthor{huisman2001accurate}.
Additionally, our normalization improves the \ac{auc} score related to
the detection of \ac{cap} in \ac{cg}, also known to be the most
challenging cases in the diagnosis of \ac{cap}.

However, using the entire enhanced signal in conjunction with the
normalization is limited by one drawback: the training time of the
\ac{rf} classifier increases; instead of using $3$ to $5$ features, so the
feature space becomes a $40$ dimensional space. The extraction of the
semi-quantitative parameters leads to a comparable classification performance ---
$0.655 \pm 0.108$ vs. $0.666 \pm 0.154$ --- and should be chosen as the alternative method to
consider if the number of feature in the classification is
critical. Despite this fact, the benefit of our proposed normalization method
has been shown for both methods.

Nevertheless, this study is performed on a small cohort of patients using a single \ac{mri} machine.
Generalizing the results of this study onto a larger dataset acquired
from different commercial systems needs to be considered to study the robustness of the proposed approach.
Additionally, our method being non-parametric should provide an
adequate framework to process \ac{dce} data from different
institutions, scanners with different settings.

### 2017-01-27 05_conclusions/conclusions.tex
Address the comments of robert
make english correction

\section{Discussions}

>>>>

\section{Conclusions and future works}\label{sec:conclusions}

In this work, we presented a new method for normalizing/standardizing \ac{dce}-\ac{mri} data.
This method is designed to reduce the inter-patient variations that occur during data acquisition.
A graph-based approach was used to correct intensity offset along with a model-based correction to reduce time offset and intensity scaling.
We show the benefit of our normalization method prior to extract quantitative and semi-quantitative features, with a significant improvement of the classification performance.
Nevertheless, we also show that using the whole normalized \ac{dce}-\ac{mri} signal outperforms all quantitative approaches.

In future research, this normalization needs to be part of an
\ac{mpmri} \ac{cad} system in which the \ac{dce}-\ac{mri} modality needs to be combined with other complementary modalities.

### 2017-01-27 main.tex
Update the acknowledgement
make english correction

% Start the abstract
\begin{abstract}
\Ac{mpmri} plays a major role in the detection of \ac{cap} and is commonly used to the design of new \ac{cad} systems.
\ac{dce}-\ac{mri} is one of the modality regularly used in \ac{mpmri} \ac{cad} system.
Pharmocokinetric parameters are extracted from the \ac{dce}-\ac{mri} sequences and later used to discriminate cancerous from healthy tissue.
However, some inter-patients variations occur during the data acquisition leading to estimation errors of the pharmacokinetic parameters.
Therefore, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduces the inter-patient variability of the data.
The benefit and simplicity of our approach is shown by detecting cancerous voxel from healthy voxel using the normalized enhanced \ac{dce}-\ac{mri} signal and comparing with the state-of-the-art quantitative and semi-quantitative methods.
Additionally, we show that using this normalization approach in conjunction with the quantitative methods improves the classification performance of most of the models.
The best classification performance to detect \ac{cap} is obtained using the whole \ac{dce}-\ac{mri} normalized signal, reaching an \ac{auc} of $0.666\ (\pm 0.154)$ and outperforming the quantitative approaches.
\end{abstract}

@@

% Reset all the acronyms
\acresetall
\linenumbers
% Include now the different files
\input{01_introduction/introduction}
\input{02_methods/methods}
\input{03_experiments/experiments}
\input{04_discussions/discussions}
\input{05_conclusions/conclusions}
\input{06_appendix/appendix}

>>>>

% Start the abstract
\begin{abstract}
\Ac{mpmri} plays a major role in the detection of \ac{cap} and is commonly used in the design of new \ac{cad} systems.
\ac{dce}-\ac{mri} is one of the modalities regularly used in \ac{mpmri} \ac{cad} systems.
Pharmocokinetic parameters are extracted from the \ac{dce}-\ac{mri}
sequences and are later used to discriminate cancerous tissue from healthy tissue.
However, some inter-patients variations occur during the data acquisition leading to estimation errors of the pharmacokinetic parameters.
Therefore, we propose a fully automatic normalization method for \ac{dce}-\ac{mri} that reduces the inter-patients variability of the data.
The benefit and simplicity of our approach is shown by detecting
cancerous voxels from healthy voxels using a normalized enhanced
\ac{dce}-\ac{mri} signal and comparing it with state-of-the-art quantitative and semi-quantitative methods.
Additionally, we show that using this normalization approach in conjunction with the quantitative methods improves the classification performance of most of the models.
The best classification performance for detecting \ac{cap} is obtained using the whole \ac{dce}-\ac{mri} normalized signal, reaching an \ac{auc} of $0.666\ (\pm 0.154)$ and outperforming the quantitative approaches.
\end{abstract}

@@

% Reset all the acronyms
\acresetall
\linenumbers
% Include now the different files
\input{01_introduction/introduction}
\input{02_methods/methods}
\input{03_experiments/experiments}
\input{04_discussions/discussions}
\input{05_conclusions/conclusions}

\section*{Acknowledgements}
Guillaume Lema\^itre is supported by the Generalitat de Catalunya (grant nb. FI-DGR2012) and partly by the Mediterranean Office for Youth (grant nb. 2011/018/06).

\input{06_appendix/appendix}
